---
title: "Common EOFs for GCM evaluation"
author: "R.E.Benestad, Met Norway."
date: "1/4/2022"
output: pdf_document
references:
- id: benestad2021
  title: A Norwegian Approach to Downscaling
  author:
  - family: Benestad
    given: R.E.
  container-title: Geosci. Model Dev. Discuss. 
  volume: NA
  URL: https://gmd.copernicus.org/preprints/gmd-2021-176/
  DOI: 10.5194/gmd-2021-176
  issue: NA
  publisher: Copernicus
  page: NA
  type: preprint
  issued:
    year: 2021
    month: 7
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This is supporting material to empirical-statistical downscaling of temperature and precipitation over the Nordic countries. It is important to evaluate how well global climate models (GCMs) reproduce essential regional climate characteristics before using them as a basis for climate change adaptation. This analysis provides an evaluation of the predictors used to represent the large-scale conditions in downscaling, as they are simulated by the GCMs. The downscaling makes use of common empirical orthogonal functions (EOFs) to represent the predictors, as explained in @benestad2021, and hence, the evaluation needs to establish whether they are skillfully reproduced by the GCMs. 

Another objective was to compare the evaluation of two generations of GCM simulations: CMIP5 and CMIP6. We focused on the models' ability to reproduce the mean annual cycle, the interannual variability and historical trends over 1950-2020 in this case. 

More information about EOFs and common EOFs are available from <https://doi.org/10.1142/6908> and a short hands-on demonstration is available ads a YouTube <https://youtu.be/32mtHHAoq6k>.

## Set-up

The evaluation of global climate model (GCM) results based on common EOFs was implemented with the help of the R-package `esd`. The following chunch activates it and sets the working environment 

```{r set-up}
library(esd)
```

### Special functions

The following code chunk defines a number of functions used in this analysis:

```{r functions}
## Function to split the CMIP file names into model, ssp and RIPF
decipher <- function(x) {
  i <- gregexpr('_',x)[[1]]
  model <- substr(x,i[2]+1,i[3]-1)
  ssp <- substr(x,i[3]+1,i[4]-1)
  ripf <- substr(x,i[4]+1,i[4]+8)
  interval <- substr(x,i[5]+1,regexpr('.nc',x)[[1]]-1)
  return(c(model,ssp,ripf,interval))
}

## Method for calculating common EOFs of the GCMs' mean annual cycle
EOF.GCMsAC <- function(x,lon=NULL,lat=NULL,...) {
  ## This function combines the annual cycles from different GCMs and performs a
  ## common EOF on the joint data
  #print('EOF.GCMsAC')
  runs <- names(x)
  runs <- runs[-grep('era5',runs)]
  #print(runs)
  if (length(grep('era5',names(x)))>0) X <- subset(x[['era5']],is=list(lon=lon,lat=lat))
  n <- length(runs)
  for (i in 1:n) {
    #print(runs[i])
    z <- subset(x[[runs[i]]],is=list(lon=lon,lat=lat))
    X <- combine(X,z)
  }
  eof <- EOF(X,anomaly=FALSE,...)
  invisible(eof)
}

## Method for calculating common EOFs of the GCMs' annually aggregated data

EOF.GCMsAM <- function(x,lon=NULL,lat=NULL,...) {
  ## This function combines the annual means from different GCMs and performs a
  ## common EOF on the joint data
  #print('EOF.GCMsAM')
  runs <- names(x)
  runs <- runs[-grep('era5',runs)]
  #print(runs)
  if (length(grep('era5',names(x)))>0) X <- subset(x[['era5']],is=list(lon=lon,lat=lat))
  X <- regrid(X,is=list(lon=seq(min(lon(X)),max(lon(X)),by=1),lat=seq(min(lat(X)),max(lat(X)),by=1)))
  n <- length(runs); print(n)
  for (i in 49:n) {
    #print(runs[i])
    z <- subset(x[[runs[i]]],is=list(lon=lon,lat=lat))
    X <- combine(X,z)
  }
  eof <- EOF(X,anomaly=FALSE,...)
  invisible(eof)
}

## Convert the list with trend maps into a field object to speed up the analysis and to 
## enable EOFs for reducing the data volume.
list2field <- function(x,it='annual',lon=NULL,lat=NULL,plot=TRUE,...) {
  #print('list2field')
  cn <- names(x)
  if (!is.null(it)) if (is.character(it)) cn <- names(x)[grep(it,names(x))]
  n <- length(cn)
  print(cn)
  Y <- subset(x[[cn[1]]],is=list(lon=lon,lat=lat)) # GCMs
  nxy <- length(Y)
  X <- matrix(rep(NA,n*nxy),n,nxy)
  for (i in 1:n) {
    z <- regrid(subset(x[[cn[i]]],is=list(lon=lon,lat=lat)),is=Y)
    X[i,] <- c(z)
    #print(c(i,cn[i],round(mean(X[i,],na.rm=TRUE),2)))
  }
  
  X <- as.field(zoo(X,order.by=1:n),param=param,unit=unit(Y),lon=lon(Y),lat=lat(Y),
                info='trend maps: list2field (GCMsAC.R)',longname='trend coefficients')
  
  if (plot) map(subset(X,is=list(lon=c(0,40),lat=c(58,70))),FUN='sd',
                main='Multi-modal ensemble trend spread (CMIP6)',sub='1950-2020')
  #par(fig=c(0,0.3,0,0.3),new=TRUE)
  #hist(c(r),lwd=3,col=rgb(0,0,0,0.5))
  invisible(X)
}
```

## Calculations

The following chunk of code is used for initialising the calculations.

```{r settings}
par0 <- par()
plot <- TRUE
reprocess <- FALSE
outpath <- '~/R/'
lon <- c(-5,40); lat <- c(55,72)
region <- paste0(paste(lon,collapse='-'),'E.',paste(lat,collapse='-'),'N')
paths <- c('~/data/CMIP5.monthly/','~/data/CMIP6.monthly/')
run <- 'r1i1p1f1|r1i1p1'
```

### Postprocessing

The postprocessing can be a bit time-consuming and the code therefore searches for results saved previously in the specified path. It will only carry out all the prosprecessing if there is no data files with the expected file name. 

```{r postprocessing}
for (path in paths) { 
  rcps <- list.files(path=path)
  rcps <- rcps[grep("RCP45|ssp245",rcps)]
  for (rcp in rcps) { 
    print(rcp)
    for (param in c('tas', 'pr', 'psl')) {
      if (param=='pr') {
        gcmfiles <- list.files(path=paste0(path,rcp),pattern='pr_Amon',full.nams = TRUE)
        gcmfiles <- gcmfiles[grep(run,gcmfiles)]
        gcmfiles <- gcmfiles[grep('_1850-|_1860-|_1861-',gcmfiles)]
        reanalysis <- '~/data/ERA5/ERA5_tp_mon.nc'
      } else if (param=='tas') {
        gcmfiles <- list.files(path=paste0(path,rcp),pattern='tas_Amon',full.naes = TRUE)
        gcmfiles <- gcmfiles[grep(run,gcmfiles)]
        gcmfiles <- gcmfiles[grep('_1850-|_1860-|_1861-',gcmfiles)]
        reanalysis <- '~/data/ERA5/ERA5_t2m_mon.nc'
      } else if (param=='psl') {
        gcmfiles <- list.files(path=paste0(path,rcp),pattern='psl_Amon',full.naes = TRUE)
        gcmfiles <- gcmfiles[grep(run,gcmfiles)]
        gcmfiles <- gcmfiles[grep('_1850-|_1860-|_1861-',gcmfiles)]
        reanalysis <- '~/data/ERA5/ERA5_slp_mon.nc'
      }
      
      output <- paste0('GCMsACAM.',param,'.',tolower(rcp),'.',region,'.rda')
      #   eof.output <- paste0('eof.',output)
      
      print(output)
      ##gcmfiles <- gcmfiles[1:3] ## Testing
      
      if (!file.exists(output) | (reprocess)) {
        print(gcmfiles); i <- 0
        Y <- list(); Z <- list(); W <- list(); AM <- list()
        
        print(paste('Number of CMIP6 files to read=',length(gcmfiles)))
        for (gcmfile in gcmfiles) {
          ## Estimate the mean annual cycle
          print(gcmfile)
          X <- retrieve(gcmfile,it=c(1950,2020),lon=lon,lat=lat)
          if (param=='pr') {
            X <- X * 30 ## mm/month
            attr(X,'unit') <- 'mm/month'
            FUN='sum'
          } else FUN='mean'
          print(c(varid(X),unit(X),round(range(X,na.rm=TRUE))))
          #cim <- paste(attr(X,'model_id'),attr(X,'realization'),i,sep='.')
          cim <- paste(decipher(gcmfile)[1:3],collapse='.')
          ## Mean annual cycle
          acm <- aggregate(X,month)
          ## Mean annual variability
          acs <- aggregate(X,month,FUN='sd')
          ## Annual total
          act <- annual(X,FUN=FUN)
          AM[[cim]] <- act
          ## Past trends
          if (plot) par(par0)
          act <- map(annual(X),FUN='trend',plot=plot)
          print('Trend statistics'); print(summary(c(act)))
          actDJF <- map(annual(subset(X,it='djf'),nmin=3),FUN='trend',plot=plot)
          actMAM <- map(annual(subset(X,it='mam'),nmin=3),FUN='trend',plot=plot)
          actJJA <- map(annual(subset(X,it='jja'),nmin=3),FUN='trend',plot=plot)
          actSON <- map(annual(subset(X,it='son'),nmin=3),FUN='trend',plot=plot)
          Y[[cim]] <- acm
          Z[[cim]] <- acs
          W[[paste(cim,param,'trend.annual',sep='.')]] <- act
          W[[paste(cim,param,'trend.DJF',sep='.')]] <- actDJF
          W[[paste(cim,param,'trend.MAM',sep='.')]] <- actMAM
          W[[paste(cim,param,'trend.JJA',sep='.')]] <- actJJA
          W[[paste(cim,param,'trend.SON',sep='.')]] <- actSON
          #map(Y[[cim]])
          i <- i + 1
        }
        ## Also show the era5 reanalysis
        print('Retrieve ERA5 reanalysis also')
        X <- retrieve(reanalysis,it=c(1950,2020),lon=lon,lat=lat) 
        if (param=='pr') {
          X <- X * 30 * 1000 ## m/day -> mm/month
          attr(X,'unit') <- 'mm/month'
          FUN='sum'
        } else FUN='mean'
        acm <- aggregate(X,month)
        acs <- aggregate(X,month,FUN='sd')
        ## Annual total
        act <- annual(X,FUN=FUN)
        Y[['era5']] <- acm
        Z[['era5']] <- acs
        AM[['era5']] <- act
        act <- map(annual(X),FUN='trend',plot=plot)
        actDJF <- map(annual(subset(X,it='djf'),nmin=3),FUN='trend',plot=plot)
        actMAM <- map(annual(subset(X,it='mam'),nmin=3),FUN='trend',plot=plot)
        actJJA <- map(annual(subset(X,it='jja'),nmin=3),FUN='trend',plot=plot)
        actSON <- map(annual(subset(X,it='son'),nmin=3),FUN='trend',plot=plot)
        W[['era5.trend.annual']] <- act
        W[['era5.trend.DJF']] <- actDJF
        W[['era5.trend.MAM']] <- actMAM
        W[['era5.trend.JJA']] <- actJJA
        W[['era5.trend.SON']] <- actSON
        GCMs.AC <- Y; GCMs.std <- Z; GCMs.trend <- W; GCMs.AM <- AM
        save(GCMs.AC,GCMs.std,GCMs.trend,GCMs.AM,file=paste0(outpath,output))
      } else load(output)
    }
  } 
}
```

### Dignostics

The diagnostics involve calculating and plotting the common EOFs of the postprocessed data. 

```{r diagnostics-TAS,fig.height=7}
##--------------------------------------------------------------
## Diagnostics

for (param in c('tas', 'pr', 'psl')[1]) {
  GCMs.AC.all <- NULL; GCMs.trend.all <- NULL; GCMs.AM.all <- NULL
  for (rcp in c('rcp45','ssp245')) { 
    output <- paste0('GCMsACAM.',param,'.',rcp,'.',region,'.rda')
    load(output)
    if (is.null(GCMs.AC.all)) {GCMs.AC.all <- GCMs.AC; GCMs.AC.all$era5 <- NULL} else 
      GCMs.AC.all <- c(GCMs.AC.all,GCMs.AC)
    if (is.null(GCMs.trend.all)) {GCMs.trend.all <- GCMs.trend; GCMs.trend.all$era5 <- NULL} else 
      GCMs.trend.all <- c(GCMs.trend.all,GCMs.trend)
    if (is.null(GCMs.AM.all)) {GCMs.AM.all <- GCMs.AM; GCMs.AM.all$era5 <- NULL} else 
      GCMs.AM.all <- c(GCMs.AM.all,GCMs.AM)
  }
  GCMs.AC.all -> GCMs.AC; rm('GCMs.AC.all')
  GCMs.trend.all -> GCMs.trend; rm('GCMs.trend.all')
  GCMs.AM.all -> GCMs.AM; rm('GCMs.AM.all')
  print(names(GCMs.AC))
  
  ## Fix problems
  n <- length(GCMs.AC)
  for (i in 1:n) {
    attr(GCMs.AC[[i]],'source') <- names(GCMs.AC)[i]
    attr(GCMs.AM[[i]],'source') <- names(GCMs.AM)[i]
  }
  ## CIESM_ssp245_r1i1p1f1 seems to have units of m rather than mm
  if (param=='pr') {
    for (i in 1:n) { 
      if (max(GCMs.AC[[i]]) < 1) GCMs.AC[[i]] <- 1000*GCMs.AC[[i]]
    }
  }
  
  ## Show the CMIP5 and CMIP56 with different colours
  col <- names(GCMs.AC)
  col <- col[grep('rcp|ssp',col)]
  col[grep('rcp',col)] <- 'red'; col[grep('ssp',col)] <- 'blue'
  print(table(col))
  
  ## Use common EOFs of mean annual cycle to compare the GCMs' ability to reproduce the
  ## mean annual cycle with that seen in the ERA5 reanalysis
  ceof.ac <- EOF.GCMsAC(GCMs.AC)
  if (plot) { 
    par(par0)
    plot(ceof.ac,col=col,alpha=0.3,new=FALSE)
    par(par0)
    plot(ceof.ac,col=col,alpha=0.3,ip=2,new=FALSE)
  }
  
  ## Use common EOFs of annually aggregated data to compare theinterannual variability 
  ## simulated by GCMs with the ERA5 reanalysis
  ceof.am <- EOF.GCMsAM(GCMs.AM)
  if (plot) {
    par(par0)
    plot(ceof.am,col=col,alpha=0.3,new=FALSE)
    par(par0)
    plot(ceof.am,col=col,alpha=0.3,ip=2,new=FALSE)
  }
  
  ## Use EOFs of trend maps to assess the trends in the GCMs:
  if (plot) par(par0)
  Z <- list2field(GCMs.trend,plot=plot)
  Z <- subset(Z,is=list(lon=c(5,35),lat=c(58,72)))
  teof <- EOF(Z,anomaly=FALSE)
  if (plot) {
    par(par0)
    plot(teof,new=FALSE)
    par(par0)
    plot(teof,ip=2,new=FALSE)
  }
  
  ## Another way to evaluate is with a scatterplot of the PCs of the two leading modes:
  n <- dim(teof)[1]
  tnms <- names(GCMs.trend); tnms <- tnms[grep('annual',tnms)]
  print(tnms)
  icmip5 <- grep('rcp',tnms)
  icmip6 <- grep('ssp',tnms)
  if (plot) {
    plot(teof[,1],teof[,2],xlab='Mode 1',ylab='Mode2'); 
    points(teof[n,1],teof[n,2],pch=19,col='black',cex=1.5);
    points(teof[icmip5,1],teof[icmip5,2],col='red',lwd=2)
    points(teof[icmip6,1],teof[icmip6,2],col='blue',lwd=2)
    text(teof[-n,1],teof[-n,2],1:(n-1),cex=0.6,col='grey40',pos=1)
  }
}
```


The plots of the common EOFs show the spatial pattern (upper left), the variance (based on the eigenvalues) associated with each pattern and the principal components (PCs) representing each GCM (red curves) and the ERA5 reanalysis (black curve). 

### Surface air temperature (TAS) 

The mean seasonal cycle of the surface air temperature (TAS) can more or less be represented by one pattern explaining 98% of the variance. It indicates that the seasonal cycle in TAS is coherent over all the Nordic countries with most pronounced amplitude in the eastern part of Finland and northwestern part of Russia. While all GCMs are able to reproduce the mean seasonal cycle, some have a tendency to somewhat exaggerate its amplitude.  

The evaluation of the CMIP5 models are shown as red curves, whereas CMIP6 is shown in blue. It is apparent that the blue curves are closer to the black curve representing the ERA5 reanalysis, indicating that the CMIP6 simulations have greater  skill than the preceding CMIP5 generation. 

The second mode is far less important (1%) and represents TAS contrasts between land and sea, suggesting a lag between the two. Nevertheless, the red curves representing CMIP5 exhibit greater departures from the black curve than the blue ones for CMIP6.  

Common EOFs were also used to evaluate the GCMs' ability to reproduce the spatio-temporal covariance of the annual mean TAS over the Nordic region. The GCMs and reanalysis reproduce similar features, such as most pronounced inter-annual variations in the northeastern part of the domain. They all indicate similar magnitudes. The leading mode represents 80% of the variance. For the interannual variability, both CMIP5 and CMIP6 indicate similar amplitude and trend as ERA5. 

The second annual mean TAS mode (9% of the variance) exhibits a northwest-southeast temperature gradient. 

The standard deviation ($\sigma$) of the 1950-2020 annual mean TAS linear trend estimates suggests greatest scatter in the eastern part of the domain, and an EOF analysis of the trend patterns (carried out on the absolute values) suggests that most GCMs reproduced the observed historical trend with strongest amplitude in the northeast (98% variance for the leading mode). The end point of the PC represents the ERA5 reanalysis. The model results exhibited a scatter around the observations. The second EOF suggested that the simulated trends also varied in terms of a northwest-southeast temperature gradient. These anomalies may possibly be connected to internal variations and the position of the jetstream.


A scatter plot of the PCs of the two leading modes suggested that most of the GCM results for TAS were consistent with those for the leading mode based on ERA5 (black symbol in the scatter plot), albeit with most GCMs shifted towards lower values for the second mode (gradient TAS structure). The red symbols mark the CMIP5 and the blue CMIP6 PC weights for the trend maps. The number shown near the symbols indicate the model associated with the data point. 

### Total precipitation (TP) 

```{r diagnostics-TP,fig.height=7}
for (param in c('tas', 'pr', 'psl')[2]) {
  GCMs.AC.all <- NULL; GCMs.trend.all <- NULL; GCMs.AM.all <- NULL
  for (rcp in c('rcp45','ssp245')) { 
    output <- paste0('GCMsACAM.',param,'.',rcp,'.',region,'.rda')
    load(output)
    if (is.null(GCMs.AC.all)) {GCMs.AC.all <- GCMs.AC; GCMs.AC.all$era5 <- NULL} else 
      GCMs.AC.all <- c(GCMs.AC.all,GCMs.AC)
    if (is.null(GCMs.trend.all)) {GCMs.trend.all <- GCMs.trend; GCMs.trend.all$era5 <- NULL} else 
      GCMs.trend.all <- c(GCMs.trend.all,GCMs.trend)
    if (is.null(GCMs.AM.all)) {GCMs.AM.all <- GCMs.AM; GCMs.AM.all$era5 <- NULL} else 
      GCMs.AM.all <- c(GCMs.AM.all,GCMs.AM)
  }
  GCMs.AC.all -> GCMs.AC; rm('GCMs.AC.all')
  GCMs.trend.all -> GCMs.trend; rm('GCMs.trend.all')
  GCMs.AM.all -> GCMs.AM; rm('GCMs.AM.all')
  print(names(GCMs.AC))
  
  ## Fix problems
  n <- length(GCMs.AC)
  for (i in 1:n) {
    attr(GCMs.AC[[i]],'source') <- names(GCMs.AC)[i]
    attr(GCMs.AM[[i]],'source') <- names(GCMs.AM)[i]
  }
  ## CIESM_ssp245_r1i1p1f1 seems to have units of m rather than mm
  if (param=='pr') {
    for (i in 1:n) { 
      if (max(GCMs.AC[[i]]) < 1) GCMs.AC[[i]] <- 1000*GCMs.AC[[i]]
    }
  }
  
  ## Show the CMIP5 and CMIP56 with different colours
  col <- names(GCMs.AC)
  col <- col[grep('rcp|ssp',col)]
  col[grep('rcp',col)] <- 'red'; col[grep('ssp',col)] <- 'blue'
  print(table(col))
  
  ## Use common EOFs of mean annual cycle to compare the GCMs' ability to reproduce the
  ## mean annual cycle with that seen in the ERA5 reanalysis
  ceof.ac <- EOF.GCMsAC(GCMs.AC)
  if (plot) { 
    par(par0)
    plot(ceof.ac,col=col,alpha=0.3)
    par(par0)
    plot(ceof.ac,col=col,alpha=0.3,ip=2)
  }
  
  ## Use common EOFs of annually aggregated data to compare theinterannual variability 
  ## simulated by GCMs with the ERA5 reanalysis
  ceof.am <- EOF.GCMsAM(GCMs.AM)
  if (plot) {
    par(par0)
    plot(ceof.am,col=col,alpha=0.3)
    par(par0)
    plot(ceof.am,col=col,alpha=0.3,ip=2)
  }
  
  ## Use EOFs of trend maps to assess the trends in the GCMs:
  if (plot) par(par0)
  Z <- list2field(GCMs.trend,plot=plot)
  Z <- subset(Z,is=list(lon=c(5,35),lat=c(58,72)))
  teof <- EOF(Z,anomaly=FALSE)
  if (plot) {
    par(par0)
    plot(teof)
    par(par0)
    plot(teof,ip=2)
  }
  
  ## Another way to evaluate is with a scatterplot of the PCs of the two leading modes:
  n <- dim(teof)[1]
  tnms <- names(GCMs.trend); tnms <- tnms[grep('annual',tnms)]
  print(tnms)
  icmip5 <- grep('rcp',tnms)
  icmip6 <- grep('ssp',tnms)
  if (plot) {
    plot(teof[,1],teof[,2],xlab='Mode 1',ylab='Mode2' ); 
    points(teof[n,1],teof[n,2],pch=19,col='black',cex=1.5);
    points(teof[icmip5,1],teof[icmip5,2],col='red',lwd=2)
    points(teof[icmip6,1],teof[icmip6,2],col='blue',lwd=2)
    text(teof[-n,1],teof[-n,2],1:(n-1),cex=0.6,col='grey40',pos=1)
  }
}
```


The leading common EOF for the mean seasonal cycle of TP explains 67% of the variance and is dominated by the rainfall along the west coast of Norway. The GCMs seem to reproduce the same feature, with more pronounced rainfall in autumn and winter. The simulated amplitude varies somewhat between the GCMs, but they more or less reproduce the same covariance structure found in the ERA5 reanalysis.

The blue curves clustered more tightly around the black ERA5 results, also indicating improved skill over the CMIP5 results. 

The second mode for the mean annual cycle represented about 22% of the variance of a precipitation pattern that is more uniform over the Nordic region land, and is associated with a peak in July-September. The exact timing of this peak, which may be associated with convective activity, varies somewhat between the GCMs. However, they all approximately reproduce the features found in ERA5.

In general, the blue curves were closer to the black ERA5, however, there were also some that deviated with lower weights during summer. The red curves representing CMIP5 were more systematically on the lower side in summer than CMIP6. The second mode had a character that can be associated with convective activity, and these results may suggest that the CMIP6 simulations were more skillful at simulating these processes. 

The spatio-temporal covariance structure for the annual precipitation, a predictor for interannual  precipitation, was well captured by all GCMs. The leading mode (42% of the variance) was dominated by the hotspot along the west coast of Norway whereas the second mode (17%) captured a region stretching from southern Norway and Sweden and extends into the Baltic Sea. This second mode may represent rainfall connected with low-pressure systems traveling into the Baltic Sea. Both red and blue curves clustered around the black curve with similar amplitudes and trend in terms of the first mode and slightly reduced amplitude for the second mode. 

The spread in the trend of the 1950-2020 annual precipitation totals suggests greatest spread between the models along the west and northern coast of Norway. An EOF analysis of these trend maps reconfirmed this pattern (76% of the variance), but did not suggest features of different characters in some models, as was the case for TAS. The second mode of the annual total TP trend maps explained 8% of the variance and picked up the contrasting TP statistics along the west coast of Norway and the rest of the Nordic region. 

A scatter plot of the PCs of the two leading modes suggested that the GCM results for TP were biased compared with those based on ERA5 (black symbol in the scatter plot), albeit with most GCMs shifted towards lower values for the leading mode. Both  CMIP5 and CMIP6 were biases, except for one CMIP6 outlier. The scatter of the CMIP6 second mode points was centered more around the ERA5 results than the CMIP5 results which generally were biased towards greater values.   

## Mean sea-level pressure

```{r diagnostics-SLP,fig.height=7}
##--------------------------------------------------------------
## Diagnostics

for (param in c('tas', 'pr', 'psl')[3]) {
  GCMs.AC.all <- NULL; GCMs.trend.all <- NULL; GCMs.AM.all <- NULL
  for (rcp in c('rcp45','ssp245')) { 
    output <- paste0('GCMsACAM.',param,'.',rcp,'.',region,'.rda')
    load(output)
    if (is.null(GCMs.AC.all)) {GCMs.AC.all <- GCMs.AC; GCMs.AC.all$era5 <- NULL} else 
      GCMs.AC.all <- c(GCMs.AC.all,GCMs.AC)
    if (is.null(GCMs.trend.all)) {GCMs.trend.all <- GCMs.trend; GCMs.trend.all$era5 <- NULL} else 
      GCMs.trend.all <- c(GCMs.trend.all,GCMs.trend)
    if (is.null(GCMs.AM.all)) {GCMs.AM.all <- GCMs.AM; GCMs.AM.all$era5 <- NULL} else 
      GCMs.AM.all <- c(GCMs.AM.all,GCMs.AM)
  }
  GCMs.AC.all -> GCMs.AC; rm('GCMs.AC.all')
  GCMs.trend.all -> GCMs.trend; rm('GCMs.trend.all')
  GCMs.AM.all -> GCMs.AM; rm('GCMs.AM.all')
  print(names(GCMs.AC))
  
  ## Fix problems
  n <- length(GCMs.AC)
  for (i in 1:n) {
    attr(GCMs.AC[[i]],'source') <- names(GCMs.AC)[i]
    attr(GCMs.AM[[i]],'source') <- names(GCMs.AM)[i]
  }
  ## CIESM_ssp245_r1i1p1f1 seems to have units of m rather than mm
  if (param=='pr') {
    for (i in 1:n) { 
      if (max(GCMs.AC[[i]]) < 1) GCMs.AC[[i]] <- 1000*GCMs.AC[[i]]
    }
  }
  
  ## Show the CMIP5 and CMIP56 with different colours
  col <- names(GCMs.AC)
  col <- col[grep('rcp|ssp',col)]
  col[grep('rcp',col)] <- 'red'; col[grep('ssp',col)] <- 'blue'
  print(table(col))
  
  ## Use common EOFs of mean annual cycle to compare the GCMs' ability to reproduce the
  ## mean annual cycle with that seen in the ERA5 reanalysis
  ceof.ac <- EOF.GCMsAC(GCMs.AC)
  if (plot) { 
    par(par0)
    plot(ceof.ac,col=col,alpha=0.3)
    par(par0)
    plot(ceof.ac,col=col,alpha=0.3,ip=2)
  }
  
  ## Use common EOFs of annually aggregated data to compare theinterannual variability 
  ## simulated by GCMs with the ERA5 reanalysis
  ceof.am <- EOF.GCMsAM(GCMs.AM)
  if (plot) {
    par(par0)
    plot(ceof.am,col=col,alpha=0.3)
    par(par0)
    plot(ceof.am,col=col,alpha=0.3,ip=2)
  }
  
  ## Use EOFs of trend maps to assess the trends in the GCMs:
  if (plot) par(par0)
  Z <- list2field(GCMs.trend,plot=plot)
  Z <- subset(Z,is=list(lon=c(5,35),lat=c(58,72)))
  teof <- EOF(Z,anomaly=FALSE)
  if (plot) {
    par(par0)
    plot(teof)
    par(par0)
    plot(teof,ip=2)
  }
  
  ## Another way to evaluate is with a scatterplot of the PCs of the two leading modes:
  n <- dim(teof)[1]
  tnms <- names(GCMs.trend); tnms <- tnms[grep('annual',tnms)]
  print(tnms)
  icmip5 <- grep('rcp',tnms)
  icmip6 <- grep('ssp',tnms)
  if (plot) {
    plot(teof[,1],teof[,2],xlab='Mode 1',ylab='Mode2' ); 
    points(teof[n,1],teof[n,2],pch=19,col='black',cex=1.5);
    points(teof[icmip5,1],teof[icmip5,2],col='red',lwd=2)
    points(teof[icmip6,1],teof[icmip6,2],col='blue',lwd=2)
    text(teof[-n,1],teof[-n,2],1:(n-1),cex=0.6,col='grey40',pos=1)
  }
}
```

The evaluation of the mean annual cycle in SLP showed a pattern of a northwest-southeast gradient accounting for 92% of the variance. As with TAS and TP, the CMIP6 simulations (blue curves) followed the that of ERA5 (black) more closely than the CMIP5, indicating an improvement. All the CMIP results nevertheless reproduced the observed statistics, which can be explained by seasonal dependency of the cyclonic activity over the Norwegian Sea (low-pressures systems are most pronounced in winter), with some skill.

The second mode for the annual SLP cycle described a west-east gradient pattern associated with 5% of the variance. 

The interannual variability in SLP suggested a nexus of action over northern Sweden, and all CMIP simulations more or less reproduced its amplitude. There was little trend to discern in the leading mode that was associated with 82% of the variance. The second mode was associated with 9% representing an southwest-northeast gradient (probably a degenerate mode as it has similar value for its egenvalue as the third mode). The model simulations reproduced this pattern with similar amplitude as ERA5.

The trend analysis indicated greatest model spread in the north and least in the south. An EOF analysis of the annual SLP trend maps indicated a negative bias (90% of the variance), and the pattern bore some resemblance to the trends in TAS. The second mode (5%) had similar eigenvalue as the third, which means it's less meaningful to discuss the two separately. A scatterplot of the PCs of these EOFs suggest that ERA5 is within the scatter of data points from rthe CMIP simulations. 

